Types for R, Empirically Designed and Evaluated

### 2020-05-09 paper/main.tex
initial class discussion

%
%
%
%
\subsection{The R Programming Language}

@@

\AT{Fill in.}

@@

\AT{We have classes.}

@@

\AT{We support multiple classes on values.}

@@

\AT{Roughly: too complicated. We have rudimentary class types for S3, and an S4 type for S4.
Future work will explore further.}

>>>>

%
%
%
%
\subsection{The R Programming Language}
\label{sec:R}

@@

As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.
The {\it class} (rather, the {\it classes}, as values can have multiple!) of function arguments are used by R to {\it dispatch} the correct function call for the argument \AT{$\leftarrow$ wording}.
This is best served with an example:
\AT{Example, unless we talked about this earlier, in which case refer reader back.}

@@

The class of a value suggest a shape for the value, and we felt class to be a more succinct way of capturing struct-like types.

\AT{We used classes to stand-in for structs.
The discussion here is naturally tied with the discussion of structs, so we need to make sure to divide the discussion appropriately.}

@@

As R does, our type system supports multiple classes on values, and the type of a classful value is simply a list of all of its classes.

\AT{What more to say here? Discussing occurrences of multi-classes?}

@@

R has multiple, disparate object systems, with the language providing direct support for S3, S4, and R5.
Thus far, we've mainly discussed S3 and its dispatch mechanism, and indeed our class types roughly capture the S3 ``types'' of values.
We will not discuss R5 as it is still under development.

For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type.
We stuck to simple S4 types since doing anything more would be immensely complicated, and outside of the scope of this work.
For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
Coming up with a type system that accounts for all of these factors, and consolidates multiple object-orientation frameworks, is an interesting problem in and of itself, and we aim to extend our type system to account for this in future work.
For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
